# Appendix: Wide vs long data {.unnumbered}

## Wide and Long format

 [Wide and long format]{.aside}
 Long advantages:

  * easier to filter, process, visualize, do statistics with
  * focused on measurement ("patient ID" or equivalent is a covariate, and so is measurement type)
 
 Wide advantages:

  * groups data by a covariate ("patient ID")
  * can be easier to manage (each column one measurement type)

## Converting from wide to long:

First, let's create a wide data set. We use a neat trick here that you have
not seen before: you can use the base R function `read.table` to read data
directly from a character string, rather than a file!


[`read.table(text=...)`]{.aside}
```{r}
wide <- read.table(header=TRUE, text='
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
wide
```

This is clearly a wide format – each row corresponds not to one observation
(one measurement), but one subject. We want to convert this to long format,
where each row corresponds to one observation (and therefore, for each
subject, there are three rows).


[`pivot_longer()`]{.aside}
```{r message=FALSE, warning=FALSE}
library(tidyverse)

pivot_longer(wide, cols=control:cond2,
  names_to="condition", values_to="measurement")
```

Note that we must put quotes around `condition` and `measurement` in the
code above. These are column names, but the columns don't exist yet.

## Converting from long to wide

Here is another example data set, this time in long format:

```{r}
long <- read.table(header=TRUE, text='
 subject  sampleID sex condition measurement
       1  ID000001 M   control         7.9
       1  ID000002 M     cond1        12.3
       1  ID000003 M     cond2        10.7
       2  ID000004 F   control         6.3
       2  ID000005 F     cond1        10.6
       2  ID000006 F     cond2        11.1
       3  ID000007 F   control         9.5
       3  ID000008 F     cond1        13.1
       3  ID000009 F     cond2        13.8
')
long
```

As you see, there are three subject, each with three measurements (one
control, one for condition 1, and one for condition 2). We want to convert
it to wide format, so we expect one row per subject, in total three rows.

However, watch out. The first thing we might want to try does not give the
expected result:

[`pivot_wider()`]{.aside}
```{r}
## not what we wanted!!! Why?
pivot_wider(long, names_from=condition, values_from=measurement)
```

The problem is in the sampleID column. Given that names of the variables
are in the `condition` column, and `measurement` is in the `measurement`
column, R considers all the remaining columns to be the identifier columns.
But the column `sampleID` contains only unique values, so they must be put
in separate rows, as above. 

To fix this, we need to tell R not to use `sampleID` as an identifier
column; instead, only the subject column should be used as an identifier.
We can also throw in the `sex` column if we want to keep it, since it has
only 1 value per subject.


[`pivot_wider(id_cols=...)`]{.aside}
```{r}
## Instead: 
pivot_wider(long, id_cols=c(subject, sex),
                  names_from=condition, values_from=measurement)
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-widelong-exercise}
## Converting to long format

Convert the following files to long format:

 * `labresults_wide.csv`
 * The built-in iris data set (`data(iris)`)
 * `cars.xlsx` (tricky! hint: how do you tell which value in the long
   format belongs to which row in the wide format?)

Clean up and convert to long format (what seems to be the problem? How do
we deal with that?):

 * `mtcars_wide.csv`

:::
:::
