# Reading and Writing Files


```{r echo=FALSE}
options(width=80)
```




## Aims for today

* Reading data
* Cleaning data
* Regular expressions
* Data management

Today is a special day. If there is one thing that I would like you to
learn from this course, it is how to read and write the data.

## Reading data

### Data types

In your work you will encounter many different types of data. Most
frequently, you will work with tabular data, either as Excel files or as
comma or tab separated values (CSV and TSV files, respectively). I am sure
you have worked with such files before.

[`readr` and `readxl` packages]{.aside}
To read these files, we will use two packages: `readr` and `readxl`. The
former, `readr`, is part of the `tidyverse` package, so when you load the
tidyverse using `library(tidyverse)`, `readr` is loaded as well. The latter,
`readxl`, is a separate package that you need to install and load
separately.

[`read.table()`, `read.csv()`, `read.tsv()`]{.aside}
There are also "base R" functions `read.table`, `read.csv`,
`read.tsv` (there is no function for reading XLS[X] files in base R). These
are always available when you start R, but don't use them. The tidyverse
functions are not only faster, but also much better behaving and, which is
most important, they are safer – it is less likely to mess up your data
with them.

[tibbles]{.aside}
tidyverse functions return *tibbles*, which, as you remember from
yesterday, are a special flavor of data frames. Just to refresh your
memory, here are key facts about tibbles:

 * in most of the cases, they behave exactly like data frames
 * when you print them, they are nicer
 * tibbles have no row names
 * when you select columns using `[ , sel ]`, you *always* get a tibble,
   even if you selected only one column



[`read_table()`, `read_tsv()`, `read_csv()`, `read_delim()`, `read_xls()`, `read_xlsx()`, `read_excel()`]{.aside}

---------------------------- ------------------------------- ------------------------------- --------------------------
Data type                    Function                        Package                         Notes
---------------------------- ------------------------------- ------------------------------- --------------------------
Columns separated by spaces  `read_table()`                  `readr`/`tidyverse`             one or more
                                                                                             spaces separate 
                                                                                             each column
                                                             
TSV / TAB separated values   `read_tsv()`                    `readr`/`tidyverse`             Delimiter is tab (`\t`).
                                                             
CSV / comma separated        `read_csv()`                    `readr`/`tidyverse`             Comma separated values
                                                                                             
Any delimiter                `read_delim()`                  `readr`/`tidyverse`             Customizable
                                                             
XLS (old Excel)              `read_xls()`                    `readxl`                        Avoid using XLS files.
                             `read_excel()`                                                  From the
                                                                                             `readxl`
                                                                                             package.
                                                             
XLSX (new Excel)             `read_xlsx()`                   `readxl`                        From the `readxl` package.
                             `read_excel()`                                                  You need to
                                                                                             provide the
                                                                                             sheet number
                                                                                             you wish to
                                                                                             read. 
                                                                                             Note: returns a tibble,
                                                                                             not a data frame! 
---------------------------- ------------------------------- ------------------------------- --------------------------

As you can see, the functions we recommend to use can be used by loading
the packages `tidyverse` and `readxl`. If you haven't done that yet, please
install these packages now:


```{r eval=FALSE}
install.packages("tidyverse")
install.packages("readxl")
```

However, before we start using these functions, we need to dive into a very
important problem: where are your files?

::: {.callout-important}
## Remember!
 * For reading text files (csv, tsv etc.), use the `readr` package. This
   package is loaded automatically when you load the `tidyverse` package:
   `library(tidyverse)`. Then, use the functions `read_csv`, `read_tsv` etc.
 * For reading Excel files, use the `readxl` package: `library(readxl)`.
   Then, use the function `read_excel`.
:::

### Where is my file? Relative and absolute paths

We are coming now to a crucial problem and a source of endless frustration
for beginning R programmers: how to tell R where your file is. Fortunately,
there are many ways to deal with that, both with R and RStudio. Still, this
is a key problem and we would like you to spend some time on the following
chapter.

[Path]{.aside}
In order to access a file, a program (any program) needs what is known a
*path*. Path is a character string that tells the program how to get to the
file and looks, for example, like this on a Windows computer:
`C:/Users/johndoe/Documents/RProjects/MyFirstRProject/data.csv`, and like
this on a Mac: `/Users/johndoe/Documents/RProjects/MyFirstRProject/data.csv`.

::: {.callout-tip}
## Path separators on different systems

Most computer systems separate the directories and files in a path using a
slash (`/`). However, Windows uses a backslash (`\ `). This is quite
annoying for R users, because in R character vectors, the slash
has a special meaning. To use backslashes, you need to "escape" them by
putting another backslash in front of each backslash. So instead of
`C:\Users\johndoe\Documents`, you need to write
`C:\\Users\\johndoe\\Documents`. Alternatively, you can use the forward
slash even on a Windows system, so type `C:/Users/johndoe/Documents`. We
recommend the latter approach.

This is also why simply copying a path from the Windows Explorer to your R
script will not work in R – because the copied text contains single
backslashes.
:::


If R cannot find your file, it will return an error message. At first, you
will be seeing this error *a lot*:


```{r echo=FALSE}
read_csv <- function(x) {
  msg <- paste0("'", x, "' not found in the current working directory\n('C:/Users/johndoe/Documents/RProjects/MyFirstRProject/').")
  stop(msg, call. = FALSE)
}
```

[Error: file not found]{.aside}
```{r error=TRUE,warning=FALSE,message=FALSE}
library(tidyverse)
mydata <- read_csv("data.csv")
```

```{r echo=FALSE}
rm(read_csv)
```




[`getwd()`]{.aside}
Before we proceed, you need to understand one important thing. When you
start your RStudio and R session, the R session runs in a specific
directory. This is called the *working directory*. You can check what is
using the `getwd()` function[^setwd]:

[^setwd]: You can change it using the `setwd()` function, but avoid doing
that. This path leads to madness, trust me on that.


```{r echo=FALSE}
getwd <- function() {
  return("C:/Users/johndoe/Documents/RProjects/MyFirstRProject")
}
```


```{r}
getwd()
```

```{r echo=FALSE}
rm(getwd)
```

Of course, the result will be different on your computer. By the way, the above is
called an *absolute* path. That means that it works no matter where you are
in your system, because a program can always find the file or directory
using this path.

The easiest way to read the data is this: copy your data files to the
directory returned by `getwd()`.

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-simple-reading}
### Reading your first file

 * Check your working directory using `getwd()`
 * Load the tidyverse package using `library(tidyverse)`
 * Go to the URL [https://github.com/bihealth/RCrashcourse-book/Datasets](https://github.com/bihealth/RCrashcourse-book/Datasets)
 * Click on "iris.csv"
 * Click on the "Download raw file" button on the right side of the screen 
 * Save the file in the directory returned by `getwd()`
 * Read the file using `read_csv("iris.csv")`
:::
:::

The following code should now work without an error:

```{r eval=FALSE}
library(tidyverse)
iris_data <- read_csv("iris.csv")
```

Now the contents of the file are stored in the `iris_data` object. There
are many ways to have a look at it:

 * type `iris_data` or `print(iris_data)` in the console
 * type `View(iris_data)` in the console
 * click on the little spreadsheet icon next to the `iris_data` object in
   the Environment tab in RStudio (upper right panel)

Please make sure that the above works for you. If it does not, read the
instructions again. In RStudio, there is a "Files" tab in the lower right
panel. You should see your working directory as well as the "iris.csv"
file there.

### Reading with relative paths

OK, so far, so good. That was easy. Now comes a slightly harder part.

Saving your data files in the working directory works well if you have one
or two. However, the more files you read and write, the more cluttered your
project directory becomes. You will soon find yourself in a situation where
you have no idea which file is which.

It is generally a good idea to keep your data files in a separate
directory, or even multiple directories. 

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-simple-reading-directory}
### Reading your first file from a data directory

 * In the working directory, create a new directory called "Datasets"
 * Move the "iris.csv" file to the "Datasets" directory
 * Read the file using `read_csv("Datasets/iris.csv")`
:::
:::

[Relative path]{.aside}
The path "Datasets/iris.csv" is called a *relative* path, because it is
relative to the working directory and will not work from another location.
So why should we use relative paths? Wouldn't it be easier to use absolute
paths all the time, for example `read_csv("C:/Users/johndoe/Documents/RProjects/MyFirstRProject/Datasets/iris.csv")`?

Actually, no. The problem is that if you move your R project to another
location, of if you share it with someone else, the absolute path will no
longer work. In other words, the absolute path is not portable.

::: {.callout-important}
## Remember!

Do not use absolute paths in your code. Always use relative paths.
:::

### More on relative paths

Some times the data files are not in your R project directory. For example,
you are writing your PhD thesis. You have created a directory called "PhD",
which contains directories "manuscript", "data", "images", "R_project" and so on. You
use R to do a part of the calculations, but you want to keep the data files
in the "data" directory. How to read them?

When you type `getwd()`, you will get the path to the "R_project"
directory, something like `C:/Users/johndoe/Documents/PhD/R_project`. The
data files are in the directory `C:/Users/johndoe/Documents/PhD/data`. To
get the relative path from the R project directory to the data directory,
think about how you would navigate from one directory to another in the
Windows Explorer or Finder. You need to go up one level, to get into
"PhD", and then down again to "data".

[Up one level with `..`]{.aside}
Getting "up one level" in a path is done by using `..`. So the relative
path to the file "iris.csv" in your "data" directory is `../data/iris.csv`.

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-reading-directory-relative}
### Reading a file from a data directory using relative paths

 * In the directory that contains the working directory create a directory
   called "Data". That is, if your working directory is
   `C:/Users/johndoe/Documents/PhD/R_project`, create the directory
   `C:/Users/johndoe/Documents/PhD/Data`
 * Move the "iris.csv" file to the new "Data" directory
 * Read the file using `read_csv("../Data/iris.csv")`
:::
:::

But what about the case when your data directory is at a completely
different location? For example, on a different drive, or maybe on your
desktop? 

First, I don't recommend keeping your data files separately from the R
project directory. In general, try to put everything in one place, as part
of one structure. This structure can be complex, but it should be coherent.
If necessary, copy the data files into your R project directory.

However, sometimes it simply isn't possible. Maybe the files are huge and
you need to read them from a special drive. In this case, there are three
options.

**Using absolute paths.** Yes, I told you not to use absolute paths, but
sometimes you have no choice.

**Create shortcuts.** In all systems it is possible to create shortcuts to
your data directories (on Unix-like systems like MacOS they are called
"symbolic links"). You can put these shortcuts in your R project directory
– R will treat them as normal directories.

**Create a complex relative path.** Depending on how far "away" your data
directory is, you can create a complex relative path. For example, if your
R project directory is `C:/Users/johndoe/Documents/PhD/R_project` and your
data directory is `D:/Data`, you can use the path `../../../../Data/iris.csv`.
Unfortunately, despite being a relative path, this is not very portable
(and it is easy to lose count on the `..`'s).

### Using RStudio to import files

RStudio has a very nice feature that allows you to import files using a
point-and-click interface. When you click on a data file in the "Files" tab
(lower right panel), you will see two options: "View File" and "Import
Dataset". Choosing the latter opens a dialog window which shows a preview
of the file and allows you to construct your `read_csv`, `read_tsv` or
another command using a point-and-click interface. You also see a preview
of the expression that will be executed.

This feature is particularly useful when you are not sure about the format
of the data to import, e.g. what type of delimiter is used, how many lines
should be skipped etc. 

Then you can click on "Import" to actually run the command, however I would
recommend another approach. Clicking on "Import" runs the command directly
in the console, bypassing your script – and you should always enter the
code in your script before executing it.

Rather than clicking on "Import", click on the little Copy icon next to the
"Code preview" field, and then cancel the dialog. Paste the copied code
into your script, and then run it.

There is one more thing to modify. The code generated by RStudio often uses
absolute paths. Try to modify it to use relative paths to ensure
portability of your script.

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-opening-data}
### Reading data

 * Go to the URL [https://github.com/bihealth/RCrashcourse-book/Datasets](https://github.com/bihealth/RCrashcourse-book/Datasets)
 * Download the following files: 
   * `iris.csv`
   * `meta_data_botched.xlsx` 
   * `transcriptomics_results.csv` 
 * Alternatively, you can download the whole repository as a ZIP file and unpack it.
 * Save the files in the `Datasets/` directory in your working directory – or
   another location of your choice. From now on, I will assume that this is
   the location of your data files.
 * Read the files using the appriopriate functions. Consult the table above
   for the correct function names, or use the RStudio data import feature.
   Make sure that you are using relative paths.
:::
:::

### Reading Excel files

Reading files sometimes requires diligence. This is especially true for
Excel files – they can contain multiple sheets, tables often do not start
on the first row etc.

[Reading excel files with `readxl` package]{.aside}
The package `readxl` (which you hopefully successfully used to read the
XLSX file in the previous exercise) contains several example files. They
have been installed on your system when you installed the package. Manually
finding these example files is annoying, and that is why the `readxl`
package provides a convenience function, `readxl_example()`, that returns
the absolute path to the file (yes, I know what I said about absolute
paths; this is an exception). 

[`readxl_example()`,`read_excel()`]{.aside}
```{r}
library(readxl)
fn <- readxl_example("deaths.xls")
print(fn)
deaths <- read_excel(fn)
head(deaths)
```

If you view this file using the `head()` function (or if you use the RStudio data import feature), you
[`head()`]{.aside}
will notice that the actual data in this file starts on line 5; the first 4
lines contain the text "Lots of people simply cannot resist writing some notes
at the top of their spreadsheets or merging cells". To skip these lines, we
need to use an argument to the `read_excel` function. If you look up the
help file for the function (e.g. using `?read_excel`), you will find the
following fragment:

[skipping lines with `read_excel(skip=...)`]{.aside}
```
skip	Minimum number of rows to skip before reading anything,
        be it column names or data. Leading empty rows are 
        automatically skipped, so this is a lower bound. Ignored
        if range is given.
```

Therefore, we can modify the code above to skip the first four lines:

```{r}
deaths <- read_excel(fn, skip=4)
head(deaths)
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-day3-opening-data-with-options}
### Reading data with options

If you take a closer look at the file `deaths.xls`, you will notice that
there is some extra text at the bottom of the file as well. How can you
omit that part when reading? *Hint: there are two ways to do that. If in
doubt, look up the "examples" section of the `readxl` helpfile.*

([Solution](solutions.html#sec-day3-opening-data-with-options))
:::
:::
 

## Diagnosing and cleaning data

### Diagnosing datasets {#sec-diagnosing}

Congratulations! You are now proficient in reading data into R. Believe me
or not, this alone is an important step forward – many who try their hand
at R get stuck at this point.

However, reading data is only the first step. In most of the cases, the
data requires some treatment: cleaning, transformation, filtering etc. In
many projects that I have worked on as a bioinformatician, data import,
diagnosis and cleaning took up the majority of time spent on the project.
Unfortunately, this is necessary before any real fun with the data can
start.

[The `iris` dataset]{.aside}
Let us examine the file `iris.csv` that you have just read. The dataset
comes from a famous paper by Ronald Fisher, who used it to demonstrate
his newly developed method called linear discriminant analysis – an early
machine learning algorithm, if you will. The dataset contains measurements
(in cm) of 150 flowers of three species of irises: *Iris setosa*, *Iris versicolor*
and *Iris virginica*. The measurements are the sepal and petal length and
width, four measurements in total. Each row consists of these four
measurements, plus a column which contains the species name. 

I have doctored the file to mimick typical problems with imported data,
especially in clinical trial settings. Humans who enter data make errors,
this is normal and expected, all of us do. Before we analyse them, we need
to correct them. Before we correct them, we need to *find* them.

First, note that when you read the data using `read_csv`, the function
conveniently shows what types of data were assigned to each column:

```{r}
iris_data <- read_csv("Datasets/iris.csv")
```

This should already ring some alarm bells. First, it looks like the columns
are not consistently named: we have `Sepal Length` and `Petal?Length`, and
also `Petal.Width`.

[`class()`]{.aside}
Secondly, we would expect that the measurements are imported as numbers.
However, both sepal and petal lengths are imported as characters. We can
confirm this by using the `class` function:

```{r}
class(iris_data[["Sepal Length"]])
class(iris_data[["Sepal Width"]])
class(iris_data[["Petal?Length"]])
class(iris_data[["Petal.Width"]])
```

::: {.callout-tip}
## `class` and `typeof`

The `class` function returns the class of an object, which is a higher-level
classification of the object. An object can have multiple classes. The
`typeof` function returns the internal storage type of the object, which is a lower
level classification. For example, both tibbles and data frames have the
type `list`, but their classes are different.
Another example:i if `mtx` is a matrix of numbers, `typeof(mtx)` is
`double`, and `class(mtx)` is `matrix`.
:::


Note that instead of using `iris_data$Sepal Length` (which will not work,
because there is a space in the column name), we use the double bracket
notation. An alternative would be to use quotes: `iris_data$'Sepal Length'`. 
This a reason why want to avoid spaces and special characters in column names
(in a moment we will show you how to standardize column names). If you use
tab-completion with RStudio, the quotes will be inserted automatically.

[`summary()`]{.aside}
We can also use the `summary` function on the whole dataset:

```{r}
summary(iris_data)
```

Ops, another problem. Did you notice the maximum value for sepal widths? It
is 36 cm. This cannot be, way too large for the flowers. Also, both the
mean and median are around 3 cm, so this must be either an outlier or a
clerical error.

::: {.callout-tip}
## The `summary()` functions

Under the hood, there is no single `summary()` function. Instead, different
classes can have different types of summaries. Whenever you produce a
result, always try the summary function with it.
:::

[`glimpse()`]{.aside}
Alternatively, we can use the tidyverse `glimpse` function. Rather then
providing a summary, this function uses a terse format to show the data
type for each column and first few values:

```{r}
glimpse(iris_data)
```

[`str()`]{.aside}
Another way how we can diagnose the dataset is to use the `str` function.
This provides a more detailed summary for each column:

```{r}
str(iris_data)
```

### Using `skim()` to diagnose datasets {#sec-skim}

[`skim()`]{.aside}
Another option to diagnose the dataset is to use the `skim` function from
the `skimr` package, which you installed (hopefully) yesterday. The `skim`
package provides a more detailed summary of the dataset, including the
number of missing values, the number of unique values and summary
statistics. To use it, you need first to load the package:


```{r message=FALSE}
library(skimr)
```

This `skim()` function shows more than just the `str` function. In addition
to column types, for character and factor columns it shows the unique
values, ordered by their frequency. For numerical values, it shows their
non-parametric summary statistics (range, median, quartiles):


```{r}
skim(iris_data)
```


[`unique()`]{.aside}
Hm, when you look at the output you might notice one more thing: for the
last column, "Species", `skim()` shows that there are six unique
values (take a look at the `n_unique` column). However, how can that be? We know that there are only three species
in this dataset. We can check this using the `unique` function:

```{r}
unique(iris_data[["Species"]])
```

There it is. Looks like whoever typed the data[^whoever], sometimes used a lower-case
species designation, and sometimes upper-case. However, for R (and for
computers in general), "versicolor" and "Versicolor" are two different things.

[^whoever]: I hate to say it, it was probably me.


[`table()`]{.aside}
More information can be gained using the `table()` function:

```{r}
table(iris_data[["Species"]])
```

Here we not only see the unique values, but also their frequency. The `table()`
function is very useful for any categorical data, as you will see later.

::: {.callout-important}
## Lower and upper case

For computers in general and R in particular, "lowercase" and "Uppercase"
are two different things. Variables `a` and `A` are different, as are
`versicolor` and `Versicolor`. This is called *case sensitivity*.
:::

### Checking individual values

We have seen before that some measurement columns were imported as
character vectors, while others were correctly imported as numbers. To
understand why, we need to have a closer look at the individual columns.

One of the columns that has been imported as a character vector is the
"Sepal Length". We will extract the column as a vector and then try to
manually convert it to a number:

[`as.numeric()`]{.aside}
```{r}
sepal_length <- iris_data[["Sepal Length"]]
as.numeric(sepal_length)
```

Two things to note: first, R issued a warning, "NAs introduced by
coercion". That means that some values could not be converted to numbers,
and were replaced by the special value `NA`[^NA]. Second, *there was no
error*. R happily allowed you to create a vector of numbers from a vector
even though there were some problems.

[^NA]: See Day 1 for more information on `NA` values.

When you look at the output, you will easily spot the `NA` values. You
could even figure out at which positions in the vector they occur. However,
this is not the way to go – what if you have not 150 values, but 150,000?
[`is.na()`]{.aside}
Instead, use the `is.na` function:

```{r}
# convert to numbers
sepal_length_n <- as.numeric(sepal_length)

# how many are NA?
sum(is.na(sepal_length_n))

# which are NA?
which(is.na(sepal_length_n))

# show the values
sepal_length[is.na(sepal_length_n)]
```

One more problem remains: the value "36" in the sepal width column. We can
check which values are greater than 10 with a similar approach:

```{r}
sepal_width <- iris_data[["Sepal Width"]]

# how many are greater than 10?
sum(sepal_width > 10)

# which are greater than 10?
which(sepal_width > 10)

# show the values
sepal_width[sepal_width > 10]
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-petal_length}
### Petal lengths

Repeat the steps above for the `Petal length` column.
:::
:::


### Diagnosing datasets: a checklist

Whenever you read a data set, you should check the following things:

 **Column names.** Are they consistent? Are they easy to type (no
 spaces, no special characters)? Are they in the correct order? Are they
 what you expect them to be? Do you understand what each column *is*?

 **Data types.** Are the columns of the correct type (e.g.
 numerical columns imported as numbers)?

 **Categorical variables.** Do you see what you expect? Is the
 number of categories correct? Do you understand what each category *is*?

 **Numerical variables.** Do the summary statistics make sense? Are
 there any outliers? Missing values? Were the variables imported
 incorrectly as characters? Why?

 **Missing values.** Are there any missing values? Why? Are they important?
 How to deal with them?

::: {.callout-tip}
## Checklist for importing data

 * Column names
 * Data types
 * Categorical variables
 * Numerical variables
 * Missing values
:::

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-day3-diagnosing}
### Botched metadata

Load the file `meta_data_botched.xlsx` using the `readxl` package. Diagnose
the problems.

([Solution](solutions.html#sec-day3-diagnosing))
:::
:::

## Mending the data

### Correcting column names

Column names should contain only letters, numbers and underscores. While
all other characters are also allowed (so a column name can be `豚肉 😀 "temp."? (°C)`), it
is *way* easier to work with columns with simple, standardized names.

You already know one way of dealing with this – simply assign new names to the columns:

```{r}
colnames(iris_data) <- c("sepal_length", "sepal_width", "petal_length",
                      "petal_width", "species")
colnames(iris_data)
```

[`janitor` package]{.aside}
However, there is a quicker way. The `janitor` package contains a function
called `clean_names` that does exactly that, automatically and for all
[`clean_names()`]{.aside}
columns. Install it with `install.packages("janitor")` if you haven't
installed it yet.

```{r message=FALSE}
library(janitor)
iris_data <- clean_names(iris_data)
colnames(iris_data)
```

My advice is to use `clean_names` on every dataset you import.

### Correcting outliers


```{r echo=FALSE}
# function for joining strings
# with commas and "and"
pand <- function(x) {

  if(length(x) == 0) {
    return("")
  }

  if(length(x) == 1) {
    return(x)
  }

  x1 <- paste(x[-length(x)], collapse=", ")
  x2 <- x[length(x)]
  return(paste(x1, "and", x2))
}
```


In the column with sepal widths, which now has the name `sepal_width`, we
found two values that probably lack the decimal point: 
`r pand(iris_data$sepal_width[iris_data$sepal_width > 10])`. 
They are at positions
`r pand(which(iris_data$sepal_width > 10))`, respectively.

It is tempting to simply substitute the data in the original data frame:

```{r eval=FALSE}
# works, but don't do it
iris_data$sepal_width[42] <- 2.3
```

Unfortunately, this is not a good idea, because this solution is a bit like
manual editing in a spreadsheet. 
There is a better way. First, **generalize the issue**. What seems to be
the problem? The values that are larger than 10 are probably missing a
decimal point. The simplest solution is to divide them by 10. Using logical
vectors, we can do that for *all* the values that are larger than 10.

```{r}
# numbers that are missing a decimal point
too_large <- iris_data$sepal_width > 10

iris_data$sepal_width[too_large] <- iris_data$sepal_width[too_large] / 10
```

This a bit hard to read, so we make it more explicit:

```{r}
# numbers that are missing a decimal point
sepal_width <- iris_data$sepal_width
too_large <- sepal_width > 10

sepal_width[too_large] <- sepal_width[too_large] / 10
sepal_width[too_large]
```

Looks good. Finally, we assign the corrected values back to the data frame:

```{r}
iris_data$sepal_width <- sepal_width
```

As the last step, we need to check whether our approach worked for all
[`any()`]{.aside}
values in the column. For this, we will use the `any()` function, which
returns `TRUE` if any of the values in a logical vector is `TRUE`:

```{r}
any(iris_data$sepal_width > 10)
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-potential-problem}
Can you spot what the potential problem with this approach is? Hint: 
what would happen if the original value was 1.10? Can you think of a better
approach? Is it actually possible to make sure that our guess is correct?
:::
:::

### Correcting categorical data

In the column "species", we found that there are two different values for
the same species. There are many ways we can handle it. Obviously, we could
manually assign the correct labels, but that would be a lot of work.

A better way would be to somehow automatically convert all the values. In
our case, the problem is quite simple: it is sufficient to put all the
values in uniform lower case, so that `Versicolor` becomes `versicolor`.

[`tolower()`]{.aside}
We can do that in one line of code thanks to the `tolower` function:

```{r}
iris_data$species <- tolower(iris_data$species)
```

[`toupper()`]{.aside}
There is also a corresponding function, `toupper`, that converts all the
letters to upper case.

Remember to check that your changes were successful:

```{r}
table(iris_data$species)
```

This was a simple case, but what if we had more complex typos? For example,
"i.versicolor", "vers.", "V" etc. In such cases, you would need to use
a more sophisticated approach. This approach involves *regular expressions*,
and will also help us to correct the sepal and petal lengths.

### Incorrectly imported numeric data

```{r echo=FALSE,warning=FALSE}
sl_prob <- is.na(as.numeric(iris_data$sepal_length))
pl_prob <- is.na(as.numeric(iris_data$petal_length))
```

We have two columns which were imported as characters, but should be
numbers. Their brand-new names are `sepal_length` and `petal_length`. All
in all, there are `r sum(sl_prob)` problems in the sepal length column and
`r sum(pl_prob)` in the petal length column, as shown in the table below:


```{r echo=FALSE,warning=FALSE,message=FALSE}
library(pander)
.x <- "sepal_length"
map_dfr(c("sepal_length", "petal_length"), ~ {
  val <- iris_data[[.x]]
  sel <- is.na(as.numeric(val))
  corrected <- val |>
    str_replace(",", ".") |>
    str_replace("[^[:digit:].]", "") |>
    as.numeric()
  tibble(
    Column = .x,
    Position = which(sel),
    Original = paste0('`', val[sel], '`'),
    Correct = corrected[sel])
}) |> pander()
```

The last column in the table above is our target – this is how we would
like to correct the data. How can we do that?

::: {.callout-tip}
## Correcting values

Of course, without additional information we can only *guess* that `5,6`
should be `5.6`. Maybe there were two measurements, 5 and 6? Maybe `5. 4`
should be `5.04`, and not `5.4`? In a real-world scenario, you would need
to consult the author(s) of the data, or check your lab book. 
:::

It is tempting to simply substitute the data in the original data frame:

```{r eval=FALSE}
# works, but don't do it
iris_data$sepal_length[70] <- 5.6
``` 

As before, that is not a good idea. If you receive another version of the
file, with added lines before the 70th, the position will change and you
will have to correct the version manually again. Also, if there are new
problems, like yet another typo in the data, you will have to spot it and
manually add a line to correct it. Again, we need to generalize the issue.

What is the problem
on line 70? A comma instead of a decimal dot – maybe someone who typed it
was used to entering numbers in a German version of Excel, where a comma is
used instead of a dot. So why not replace *all* the commas by a dot? This
should do the trick not only for line 70, but also for possible future
problems.

[`str_replace_all()`]{.aside}
We will use for that the function `str_replace_all` from
tidyverse[^stringr]. Before we do that, however, we will make a copy of the
column. That will allow us to easily see the changes will make, check that
everything is OK and only then replace the column in the data frame. Later
on, you will learn more efficient ways of doing this.

[^stringr]: In base R, you can use the `gsub` function. However,
it has a slightly different syntax, which makes it less convenient
when you start using pipes tomorrow.

```{r}
# make a copy
sepal_length <- iris_data$sepal_length

# record the problematic places
problems <- is.na(as.numeric(sepal_length))
sepal_length[problems]

# replace the comma 
sepal_length <- str_replace_all(sepal_length, ",", ".")

# what was the result?
sepal_length[problems]
```

It worked!
The `str_replace_all` is a general search-and-replace function. You can
always use it with character vectors to replace stuff.

However, we still have the problem with `> 7.7`. You will find such values
quite often in clinical settings, along with the counterpart `< 7.7`. These
may indicate that the measurement was out of range of an instrument.
However, we cannot usually treat them as missing values, because they
contain some information. Depending on whether or not we want to keep this
information, we could either replace them by NAs, or decide to keep the
value as is (in this case, change `> 7.7` to `7.7`).

In the latter case, we can use the `str_replace_all` function again:

```{r}
sepal_length <- str_replace_all(sepal_length, "> ", "")
```

In any case, the last point that remains is to convert the vector to
numbers and assign it to the column holding sepal lengths. We do it in one
go and also check if everything is OK:

```{r}
# finalize
iris_data$sepal_length <- as.numeric(sepal_length)

# check whether the column is numeric
is.numeric(iris_data$sepal_length)

# check whether our problems are gone
iris_data$sepal_length[problems]

# check whether there are any NA's
any(is.na(iris_data$sepal_length))
```

[`as.numeric()`]{.aside}
The new function, `is.numeric()`, checks whether the column `sepal_length` is
indeed numeric. Finally, we make sure that no NA's were produced in the
conversion.

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-str_replace}

Find the problems in the following vector and correct them:

```r
vec <- c(" 5", "5,6", "5.7", "5. 4", "> 5.0", "6.O")
```
:::
:::

## Regular expressions

### Introduction to regular expressions

[Regular expressions]{.aside}
Regular expressions, regexps for short, are for finding certain patterns in
the data and handling them automatically. They are quite old (going back at
least to 1960's) and have remained virtually the same for many decades. You
will find them in almost all programming languages, and while details may
differ, the principles are usually the same. The regexps in R are very
similar to regular expressions in other programming languages such as
Python. As you may guess, they have a different syntax then R – they are
something else altogether.

We will not go in depth in regular expressions in R here. Here, just a
short primer.

[`str_detect()`]{.aside}
Before we start, let us introduce a new function called `str_detect()`[^grep]. 
The function is available as soon as you load the `tidyverse` package.
This
function is used to search for a pattern in a character vector. For every
element that matches the pattern, it will return a `TRUE`, otherwise a
`FALSE`. Say, we have a vector (or data frame column) containing sample
names, and we want to find all the controls.

[^grep]: In base R, there is the `grep()` function. It is very similar, but
has a different syntax. We will stick to `str_detect()` for now.

```{r}
samples <- c("ko_1_ctrl", "ko_2_ctrl", "ko_1_treat", "ko_2_treat",
            "wt_1_ctrl", "wt_2_ctrl", "wt_1_treat", "wt_2_treat")
str_detect(samples, "ctrl")
samples[str_detect(samples, "ctrl")]
```

In case that someone did not pay attention to lower or upper case, we can
tell str_detect to ignore the case:

```{r}
samples <- c("ko_1_ctrl", "ko_2_CTRL", "ko_1_treat", "ko_2_treat",
            "wt_1_CTRL", "wt_2_ctrl", "wt_1_treat", "wt_2_treat")
# this does not work
str_detect(samples, "ctrl")
# but this does
str_detect(samples, regex("ctrl", ignore_case=TRUE))
```

Here we are looking for a literal string "ctrl". However, we can use
regular expressions to search for more complex patterns. What if the sample
names are more hap-hazard, like `ko_1_control`, `ko_2_kontrol`,
`ko_1_ctrl`?


```{r}
samples <- c("ko_1_control", "ko_2_ctrl", "ko_1_trt", "ko_2_treatment",
            "wt_1_Control", "wt_2_kontrol", "wt_1_Trtmt", "wt_2_treated")
```

OK, so we got `control`, `ctrl`, `Control` and `kontrol`. A pattern
emerges:

 * first letter is either `k` or `c` or `C`
 * then we might have an `on` (or not)
 * next always comes `tr`
 * then we might have an `o`
 * then we always have an `l`
 * then the word ends.

All this "mights" and "ors" and "eithers" and "always" can be encoded in a
regular expression:

|Pattern| Explanation|
|-------|------------|
|`[kcC]` | must have one of the letters `k`, `c` or `C`|
|`o?` | zero or one `o` (i.e., "there might be an o")|
|`n?` | zero or one `n` (i.e., "there might be an n")|
|`tr` | must have literally the string `tr`|
|`o?` | zero or one `o` (i.e., "there might be an o")|
|`l` | must have literally the string `l`|
|`$` | here must be the end of the string|

Taken together, this gives us the regular expression `[kcC]o?n?tro?l$`.
This looks weird, but has the magic power of matching all the control-like
strings in our vector. Let us try it:

```{r}
str_detect(samples, "[kcC]o?n?tro?l$")
samples[str_detect(samples, "[kcC]o?n?tro?l$")]
```

This works, but... what does that all mean?

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-regex1}
### Quick question

Look at the table above.
Can you think of other patterns that would match this regular expression? 
Would `kotrl` work? What about `cotro`?
:::
:::

### How regexp works

[regexp rules]{.aside}
A regular expression is a sequence of characters – a string. However, some
of these strings have a special meaning. Here is a short table, we will
discuss these in more detail in a moment:

|Character| Meaning|
|---------|--------|
|`a` | a literal character `a`|
|`.` | any character|
|`[abc]` | one of the characters `a`, `b` or `c`|
|`[^abc]` | any character *except* `a`, `b` or `c`|
|`[a-z]` | any character from `a` to `z`|
|`[0-9]` | any digit|

These are the so-called "atoms". They stand for something that we want to
match. In the simplest case (like the `a` above), they stand for themselves.
Also, did you notice the square brackets? They have *nothing* to do with
how we use square brackets in R.

In addition we can specify how many times we want to match the atom:

|Quantifier| Meaning|
|----------|--------|
|`?` | zero or one|
|`*` | zero or more|
|`+` | at least one|

Then, we can "anchor" the pattern in our string:

|Anchor| Meaning|
|------|--------|
|`^` | start of the string|
|`$` | end of the string|

Let us put that together. Say we have the following strings:

```{r}
strings <- c("a", "ab", "ac", "ad", "bc", "abc", "bac", "cab")
```

To match these which contain an `a`, we can use the regular expression `a`:

```{r}
strings[str_detect(strings, "a")]
```

To match these which *start* with an `a`, we can use the regular expression `^a`:

```{r} 
strings[str_detect(strings, "^a")]
```

To find these which start with an `a` followed by `b` or `c`, but not `d`, we
can use the square brackets:

```{r}
strings[str_detect(strings, "^a[bc]")]
```

OK, that was a lot. Take some time to digest it. Regulare expressions are
what you call in German *gewöhnungsbedürftig* – they require some getting
used to, but they are worth it.


### Usage of regular expressions in R

Most commonly you will use regular expressions in R for two things:

 * Searching for something in a character vector (like we did above)
 * Replacing something in a character vector, for example using `str_replace_all`

What we did not tell you when we introduced `str_replace_all` is that it
actually, its second argument is a regular expression. Therefore, we can
use it to find a pattern and replace it with something else. For example,
we can unify the following messed up vector denoting the gender of patients:


```{r}
gender <- c("m", "f", "m", "w", "frau",
            "female", "other", "male", 
            "männlich", "x", "weiblich")
```

German and English are mixed here, but we can see a pattern: if the strings
starts with `m`, it is male, if it starts with either `f` or `w`, it is
female. And in remaining cases it is "other". We can clean up this mess
with just three lines of code:


```{r}
gender <- str_replace(gender, "^m.*", "male")
gender <- str_replace(gender, "^[fw].*", "female")
gender <- str_replace(gender, "^[^mfw].*", "other")
```

Note that the `^` character has two different meanings on line 3 above. As
the first character of a regular expression, it anchors it to the beginning
of the string. However, inside the square brackets, it negates the
selection. So, `[^mfw]` means "any character except `m`, `f` or `w`".

The `.*` idiom is a common one in regular expressions. It means "zero or
more of any character". So, `^m.*` matches both the string "m" and the
string "male". And because we used `^`, it will only match the strings
where m is the first letter (so it does not match "female").

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-regex1}
### Quick question

What would happen if we ommited the `^` at the beginning of the strings above? 
For example, if we used `[^mfw].*` instead of `^[^mfw].*`?
Think first, then try it out.
:::
:::

But wait, if some characters have a special meaning, how can we replace
them? For example, how can we replace a dot in a string? The following will
not work as intended:

```{r}
vec <- c("5.6", "5.7", "5.8")
str_replace_all(vec, ".", ",")
```

Since the character `.` means "any character", *every character* will be
replaced by a comma in the example above. In order to search or replace
[escaping characters in a regexp]{.aside}
special characters, we must *escape* them – which in R means putting two
backslashes[^bck] in front of them in the regular expression.

[^bck]: In R, it is two backslashes. In other programming languages, it
is usually a single backslash.

```{r}
vec <- c("5.6", "5.7", "5.8")
str_replace_all(vec, "\\.", ",")
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-regex2}

 * Use `str_replace_all()` to make the following uniform: `c("male", "Male ", "M", "F", "female", " Female")`
 * Using `str_replace_all()` and `toupper()`, clean up the gene names such
   that they conform to the HGNC (all capital
   letters, no spaces, no dashes): `c("ankrd22", "ifng", "Nf-kb", " Cxcl 5", "CCL 6.", "ANK.r.d. 12")`
 * What regular expression matches all of the ankyrin repeat genes (but not other genes) in the following vector: `c("ANKRD22", "ank.rep.d. 12", "ANKRD-33", "ankrd23", "ANKEN", "MAPK", "ifng-1", "ANKA-REP-6")`? Ankyrin repeat domain genes are the first 4 in the vector.
:::
:::
 

### Correcting columns in `iris_data` with regular expressions

Let us now turn to another column in the data frame, the petal lengths.
Using the approach we have just learned, we can find the problematic
values:

```{r}
petal_length <- iris_data$petal_length
problems <- is.na(as.numeric(petal_length))
which(problems)
petal_length[problems]
```

We *could* use for example
`str_replace_all(sepal_length, "k", "")`, to change `5k.1` to `5.1`, but
that would not be a general solution. What if it is `j` in the new data? We
should actually remove *everything* that is not a number and not a decimal
dot. To this end, we will use the regular expressions.

We need to remove anything *but* numbers and decimal dots. We can use the
square brackets for that:

```{r}
petal_length <- str_replace_all(petal_length, "[^0-9.]", "")
```

The `0-9` means anything between 0 and 9, and the `.` is a literal dot. As
you can see, you don't have to escape the dot if it is already in the
[`[^0-9]`]{.aside}
square brackets. The `^` inside the square brackets negates the selection,
so `[^0-9.]` means exactly what we wanted.


```{r}
# check the problems
petal_length[problems]

# convert to numbers
petal_length <- as.numeric(petal_length)

# check for remaining NA's
any(is.na(petal_length))

# assign
iris_data$petal_length <- as.numeric(petal_length)
```

Done! The iris data set is now clean.


```{r}
skim(iris_data)
```

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-day3-correcting}
## Correcting metadata

In @exr-day3-diagnosing, you have diagnosed the file
`meta_data_botched.xlsx`. Now go ahead and correct the problems you have
found.
:::
:::


### Just a quick look {#sec-quick-look}

The newly cleaned up data set begs for a quick look. We already created a
simple plot before using the `plot()` function, and today I will show you
how to use the `ggplot2` package to create somewhat nicer plots. In this
case, it is rather trivial.
[`ggplot2`]{.aside}

```{r}
library(ggplot2)
ggplot(iris_data, aes(x=sepal_length, y=sepal_width, color=species)) +
  geom_point()
```

As you can see, the `ggplot` function takes two arguments: the data frame
and something which is calles *aesthetics* and which is produced by the
`aes()` function. The `aes()` function serves as a kind of link between the
data and the plot, showing how what should be plotted. In this case, we
tell `aes()` that the x-axis should be the sepal length, the y-axis the
sepal width, and the color should be determined by the species.

Then, a magical thing happens. Whatever is produced by `ggplot()` is not a
number, but nonetheless can be added with a `+` sign to other stuff. In
this case, we add a `geom_point()` function, which tells `ggplot` to plot
points at the x, y coordinates (there are many others, and you will learn
about them the day after tomorrow).

There, all done. See how *I. setosa* is different from the other two?

## Writing data

### Keep your data organized

Writing data is *way* easier than reading it. However, you always have to
come up with a name for the file, and I would like you to follow a few
points.

 * Never overwrite the original file (we will discuss it in more detail shortly).
 * Create an output directory and save your files only there, do not mix
   the original and the output files.
 * Use a meaningful name, so not `output`, but `iris_cleaned`. No spaces or
   special characters (use an underscore `_` instead).
 * Thirdly, always version your files: add a date, your initials, a version number and a date,
   e.g. `iris_cleaned_2021-09-01_JW_v1.csv`.
 * Do not use the old Excel XLS format (file extension `.xls`). Use the
   newer XLSX format (file extension `.xlsx`).

This latter point warrants an explanation. There are two main problems with
that format. Firstly, it is less portable then XLSX (the new one) – so many
programs can read XLSX, but not XLS (or, which is worse, they can, but read
it incorrectly). Secondly, the number of rows and columns in an XLS file is
severly limited (65536 rows and 256 columns). These limits are easily
reached in modern bioinformatic datasets.

### Functions to use for writing data

[`write_tsv()`, `write_csv()`,`write_xlsx()`]{.aside}

---------------------------- ------------------------------- ------------------------------- --------------------------
Data type                    Function                        Package                         Notes
---------------------------- ------------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `write_tsv()`                   `readr`                         No rownames!
                          
                                                                                             
CSV / comma separated        `write_csv()`                   `readr`                         No rownames!

XLS (old Excel)                                                                              Just don't use it. No, 
                                                                                             seriously, don't.
                                                                                             
XLSX (new Excel)             `write_xlsx()`                  `writexl`                       No rownames!
---------------------------- ------------------------------- ------------------------------- --------------------------

Just as in the case of reading data, there are several functions to write
data. There are also functions in the base R that can be used for writing
(e.g. `write.csv`), but they are not recommended.
We advise you to only use the `readr` package for writing TSV and CSV
files, and the `writexl` package for writing XLSX files.

However, keep in mind that row names are not exported with these packages.
That is why we do not use row names in this course.

::: {.callout-caution icon=false appearance="simple"}
::: {#exr-writing-data}
### Writing data

 * In your project directory, create the directory "Data_clean" (if you
   want, you can use the R function `dir.create()` for that).
 * Write the cleaned `iris_data` to a new file in the "Data_clean" directory.
   Use the `write_csv` function from the `readr` package. Use a meaningful
   name, version it and use a date.
:::
:::

## Principles of data management

### Keeping originals

As every day, we end the day with some food for thought. Today, we would
like to introduce a few important principles that you should follow in your
everyday work with data – not only in R, but in general.

The first rule is: **always keep the originals**. Whether you have received
the data from someone, extracted them from a database such as RedCap or
created it yourself, once you start working with R you should freeze your
data, version it and not touch it anymore.

Let me reiterate: **do not touch the original data**. If you absolutely
*have* to edit it by hand, make a copy and add a suffix like `_edited` or
your initials to the file. However, it is better to perform all editing
operations in R.

**Versioning.** Always include a date, and possibly your initials and
version in the file name. If you ever submitted a manuscript of a
scientific paper, you know how quickly versions can accumulate. It is
absolutely crucial to know which version of the data was used for which
analysis.

**Data editing.** Every edit operation on your data should be somehow
recorded. This is not different from a lab notebook, where you write down
every step of your experiment. When working with a spreadsheet this is hard
to do, therefore...

### Use R for data editing

Optimally, your mode of operation should never involve editing data
directly in a spreadsheet. You should read your data in R and perform all
operations here, recording them in a script. This has two huge advantages:

 * **Reproducibility.** You and others will be able to trace back all your
   operations. There will be no doubt whether and which data were edited
   and how.
 * **Automation.** When you get updated data (and you will), for example
   when new samples arrived, you will be able to re-run your script (maybe
   with a few changes) and get the updated results with minimal effort.
   Otherwise, you will have to repeat every. Single. Edit. Again.

Consider the example of the `iris.csv` file. We have seen that one of the
values of sepal width seems to be missing a decimal dot. You *could* edit that in
the Excel or CSV file directly. You might even make a note of that
somewhere. However, chances are that you were wrong – maybe that *was* a
flower wiht a 36 sepal width after all. This change influences your data
and your analysis results and *must* be documented if you were to do
science. If you did it in R, you probably have entered a comment in your
script, something like this:


```{r eval=FALSE}
# There seems to be a missing decimal dot in some of the sepal widths
# Changing it to 1/10th of the value

sepal_width <- iris_data[["Sepal Width"]]
sel <- which(sepal_width > 10)
sepal_width[sel] <- sepal_width[sel] / 10

iris_data[["Sepal Width"]] <- sepal_width
```

This concisely informs the reader that a change was made, how and why.

And what happens when Ronald Fisher raises from his grave and collects
[undead Ronald Fisher]{.aside}
another 150 flowers? You will get an updated data set with 300 flowers and
import it *again*. In the best case scenario, you will find the problem
again (or remember it). Then you will have to go into Excel and make the
same edit again. And all the other edits as well. If you did it in R, you
simply use your script again.

### Working with Excel and other spreadsheets

[How to work with Excel]{.aside}
We all use Excel, LibreOffice Calc or Google Sheets. They are great tools,
no doubt about that. However, by now you can see that certain operations
make working with R harder, and some even that can mess up your data.

If possible, you should actually avoid working with Excel when working with
scientific data. Excel does certain operations automatically (depending on
the settings), so it can change your data *without leaving a trace of the
changes*. This is the precise opposite of what you want in science.

::: {.callout-tip}
## Avoid working with Excel

Excel can automatically replace strings with dates, like changing `MARCH9`
into `9th of March`. However, `MARCH9` is a valid gene name. In fact, it turns
out that a substantial proportion (about a third!) of all Excel files
containing gene names and published as supplementary materials online
contain gene names transformed to dates. Not only that, but even though
that this has been discovered many years ago and even some genes were
officialy renamed *because of Excel*, this is still a problem. And Excel is
now able to recognize dates in many languages, exacerbating the problem
[@abeysooriya2021gene]. 
:::

However, sometimes it is unavoidable to work with Excel.

**Don't's:**

 * **Using color and font to encode information**, for example marking the different
   treatments with different background color or formatting controls in
   bold. While it is possible to read this information in R, it is not
   trivial. In addition, even outside of R formatting of cells can be lost
   when the data is passed around, and this would mean that the information
   is lost.
 * **Comments in the same cells as values**, for example `50 (measured twice)`.
   This prevents the columns to be interpreted as numbers and has to be
   dealt with in R. Create a separate column for comments, even it is just
   a single comment in a row.
 * **Meta-data information in the header.** Very often, column names
   contain additional information, for example units, dates, encoding (like
   "0 for male, 1 for female"). This makes it hard to both, import the data
   and actually use that information. Create a separate sheet in your
   spreadsheet for meta data, with one row per column name.
 * **Adding header and tail information**. Just like in the example with
   `deaths.xlsx` file, additional lines in spreadsheets require you to
   carefully select what you wish to import. Avoid it if you can, use
   another sheet.
 * **More than one table on a spreadsheet**. A spreadsheet should contain a
   single table. Otherwise, just like in case of header and tail
   information, you will have to spend some time cautiously selecting the
   areas that you want to import.
 * **Merging cells**. Merged cells are a nightmare to
   work with in R. If you have to, unmerge them before importing the data.

**Do's:**

 * **Use a single table per sheet**. If you have more than one table, use
   more than one sheet.
 * **Use a single header row**. If you have more than one header row, use
   create another sheet with meta-data on the column names.
 * **Switch off automatic conversion**. Excel can automatically change
   your data in many ways. For example, it can change gene names to dates.
   It can also change the decimal separator, which can mess up your data.
   Newer versions of Excel allow to switch off this behaviour, so do it
   ([here are the instructions](https://techcommunity.microsoft.com/t5/microsoft-365-insider-blog/control-data-conversions-in-excel-for-windows-and-mac/ba-p/4215336)).
 * **Control your import.** When importing data from CSV files to TSV
   files, Excel allows you to control which fields are imported as text,
   which as dates etc. Use this feature to make sure that your data is
   correctly imported.

## Review

Things that you learned today:

 * Reading and writing
   * `read_csv`, `read_tsv`, `read_delim`, `read_xls`, `read_xlsx`
   * reading a file from working directory
   * reading a file from a different directory using relative paths
 * Paths
   * Get the current working directory with `getwd()`
   * relative and absolute paths
 * Diagnosing datasets
   * data types: `str`, `class`, `typeof`
   * summaries: `summary`, `glimpse`
   * categorical data: `unique`, `table`
   * `skim()` from the `skimr` package
   * diagnosis checklist
 * Correcting datasets
   * converting character vectors to numbers with `as.numeric()`
   * checking a vector with `is.numeric()`
   * checking a logical vector with `any()`
   * replacing values with `str_replace_all()`
 * Principles of data management
   * keeping originals
   * versioning
   * data editing
   * working with Excel
 * Other
   * using help pages to find out about functions and their parameters
