# Introduction {.unnumbered}

```{r setup,include=FALSE}
# set the default output to SVG
knitr::opts_chunk$set(dev="svg")
```


## Five days of R

I have been teaching R to biologists and medical students for many years
now. At the Core Unit for Bioinformatics at the Berlin Institute of Health,
Charité - Universitätsmedizin Berlin, we have developed a five-day, 5 hour
per day R crash course running for the last three years. This book is a
companion to that course.

This is also the reason for how the materials in the book are arranged.
Rather then discussing everything about vectors first, then everything
about matrices etc., we start with easy things, and return to them later to
build on them. I call this "helical learning"[^helical] – we spiral around the same
topics, but each time going a bit deeper, and each time you will understand
a bit more. This is also why some topics are spread between the days – by
trial and error, we have found the amount of material that can be covered
in a day of learning.

[^helical]: I got this idea from the professor Barbara Płytycz from the
Jagiellonian University, who taught me my first "helix" of the immune system.

There are two goals of this book. The first one is that 
after five days of learning R, you will be able to load, inspect,
manipulate and save data files (such as Excel tables or CSV files), make
some basic plots and perform simple statistical tests. The second goal is
that you are in a good starting position to continue learning R on your own.

In other words, this course should give you a jump start, allowing to overcome
this first big hurdle in learning R.

## Prerequisites

### Installing R and RStudio

Before you dive in to the course, we would like to ask you to install both
R and RStudio: you need R (obviously) and RStudio is a great and very
popular interface for R. Minimal installation guide:

 1. Install R. Go to [R website](https://cran.r-project.org/) and download
    the version for your operating system. Install it.
 2. Install R Studio. Go to [R Studio installation
 page](https://posit.co/download/rstudio-desktop/) and follow the
 instructions.

This should be it. However, if you are stuck or have problems with
installing R or RStudio, please search the internet – the installation and
especially related problems unfortunately depends on your operating system
and the version of the software, so it is hard to give a general advice.

Here are some other websites with somewhat more detailed instructions:

 * [Installing R on all systems](https://techvidvan.com/tutorials/install-r/)
 * [Installing R on Windows 7, 8, 10](https://medium.com/@GalarnykMichael/install-r-and-rstudio-on-windows-5f503f708027)
 * [Installing R, RStudio and configuring RStudio](https://www.stat.colostate.edu/~jah/talks_public_html/isec2020/installRStudio.html)

Potential problems:

 1. **You already had R installed on the system.** Having multiple R
 versions may be a source of problems. If you have an old version of R,
 unless you have a good reason to keep it, uninstall it.

 2. **You need to compile packages.** Packages come in two forms:
 precompiled packages and source packages. Some exotic packages might not
 have the precompiled version for your system.
 If you need to compile packages, you will have to install additional
 packages. For example, for Windows, you need to install
 [Rtools](https://cran.r-project.org/bin/windows/Rtools/) - make sure you
 download the version that corresponds to your version of R.


### Installing R packages

During the course we will use several R packages that you need to install
on your computer. On Day 2, we will discuss installing and loading
packages, and we will make a note to install the required packages when
they are needed. However, you can also install them right now, after you
have installed R and RStudio. This can be more effective – after all,
installing might take some time.

Here is the list of packages that you will have to install:

 * `tidyverse`
 * `ggplot2`
 * `skimr`
 * `pander`
 * `readxl`
 * `writexl`
 * `janitor`
 * `broom`
 * `cowplot`
 * `ggbeeswarm`
 * `pheatmap`
 * `tinytex` (only if you want to produce PDF output from Rmarkdown)

You can install them by running the following code in your RStudio:

```{r eval=FALSE}
install.packages(c("tidyverse", "ggplot2", "skimr", "pander", 
                   "readxl", "writexl", "janitor", "broom", 
                   "cowplot", "ggbeeswarm", "pheatmap", "tinytex"))
```

## The structure of this book

This book is divided into five chapters, each corresponding to one day of
the course. At the beginning of each chapter, you will find a short list of
topics for the given day.

Some parts of the book are highlighted:

**Code blocks with output:**


```{r}
# this is a comment
x <- 1 + 1
```

The numbers on the left (if present) are not part of the code – they are
just line numbers. You can copy the code by clicking on the "Copy" button
(📋)
and it will not copy these numbers.


::: {.callout-caution icon=false appearance="simple"}
::: {#exr-example}
## Example

 * This is how an exercise looks like!
 * Please do all exercises. It helps a lot.
 * Some things are learned only through exercises.

::: {.callout-caution icon=false collapse=true}
## Solution (click me!)
Some exercises have a solution which you can click to reveal.
:::

:::
:::

::: {.callout-tip}
## Useful tips

 * Some exercises have solutions in the "Solutions" chapter. If they do,
   please read the solution after you have completed the exercise – often
   there will be a comment or a hint that will help you understand the
   material.
:::

::: {.callout-important}
## Remember!

 * Run *all* code chunks in your RStudio.
 * Do *all* exercises.
 * Go through the "Review" section at the end of each chapter and make sure
   you understand *everything* on the list.
:::



Take a look at the right margin!
[New concepts are highlighted on the right margin of the book]{.aside}

And again[^again]!

[^again]: And also footnotes.

In each chapter there are several exercises. *However*, that does not mean
that you should *only* do the exercises. In fact, you should try out
*every* piece of code that is in the book. Copy it (there is a 📋 button next
to each code block that will do it for you), paste it into your RStudio and
run it. Then try to modify it and see what happens.

Exercises in this book are important. They are not only there to check if
you understood the material, but they can also introduce new concepts or
ideas. This is because this book is not only about learning R, but also
learning *how to learn about R*. So, for example, sometimes we will want
you to figure stuff on your own rather than give you a ready-made answer.

Many of the exercises have a solution provided, either inline (you have to
expand it by clicking on the "Solution" button) or in the "Solutions"
section of the book. **It is really important that you do not give up too
quickly**. Try to solve the exercise on your own, and only then look at the
solution.

Each chapter is ended by a "Review" section, which contains a list of
things that you have learned that day. It is really important that you go
through that list and make sure that you understand everything on it. Some
of the new things appeared in the exercises, so if you skipped them, you
might want to go back and do them.

If you do all that, I personally *guarantee you* that by the end of this
course you will be able to use R in your work.

## General advice

The course is a real crash course. There is a lot of material coming at you
in a very short time. You will feel overloaded and overwhelmed – this is
normal. Don't worry! It will soon get better, and in a few days you will be
able to do fairly advanced things with R.

The key is to keep playing with your R; trying out new things, breaking it.
Please go through all exercises in that book, even if they seem simple at
the first glance (some of them are tricky, others are used to smuggle in new
concepts and useful tidbits of information).

Whenever you feel you don't understand something, stop and try to figure it
out. Use internet search very liberally. Many answers can be found on sites
such as [StackOverflow](https://stackoverflow.com), 
[R-bloggers](https://www.r-bloggers.com/), or simply in the R documentation[^help]. Try out the
code you will find in these sources – just copy-paste it into your RStudio
and adapt it to your needs. Feel free to use Large Language Models (such as
GPT) – they are very good at explaining code, especially when you are
learning basic concepts.

[^help]: You can access the R documentation by typing `?function_name` in
the console. You can search for concepts using `??keyword`.

However, if you want to learn R, simply doing this course will not be
enough. You need to start using it in a real world setting. Unfortunately –
the better you already are at Excel, Word and other such tools, the harder
it will be switching to R: tasks that are a breeze in Excel will at first require
you to spend substantially more time in R. However, trust me: it pays off
in the long run. Therefore, for best results, force yourself to use R *even
if at first it is less efficient then other tools*.

Finally: programming can be fun. Most programmers I know simply enjoy doing
that. It is a satisfaction similar to building Lego models or solving
puzzles or reading a crime story – and also very much like doing
experiments. Unlike experiments, however, you can't break expensive
equipment and the results of your attempts are usually immediately visible,
no need to wait for weeks for the results. You can play, experiment, try
out various things at no expense and no risk. If you can get into this
mindset, learning R will be much, much easier. 


## Acknowledgements

I have been teaching statistics and bioinformatics for more than two
decades now. About ten years ago, I started teaching the first "R crash
course" with the goal of introducing PhD students and postdocs to R as
quickly and as painlessly as possible. The course evolved over the years,
and I had many partners and co-teachers.

In the first place, I would like to thank Carlo Pecoraro from 
[Physalia Courses](https://www.physalia-courses.org/) for the opportunity
to teach my first R crash course for Physalia.

Several times I have taught the course together with my colleague, 
Dr. Manuela Benary from the 
[Core Unit of Bioinformatics](https://cubi.bihealth.org) 
at the Berlin Institute of Health, Charité - Universitätsmedizin Berlin.
Manuela has been a great partner in teaching the course, and many ideas in
this book are actually hers.

I would also like to thank many colleagues and friends who had the patience
to go through this book and provide feedback.

Finally, I thank you, the Reader, for your interest in this book. I hope
it will be useful and that you will enjoy using R in your research.
